The Companion Mass Distribution of extremely low-mass white dwarfs

### 2014-10-12 documents/letter/ms.tex
oops figures back in

% commands
\newcommand{\given}{\,|\,}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transpose}[1]{{#1}^{\mathsf{T}}}
\newcommand{\inverse}[1]{{#1}^{-1}}
\newcommand{\Msun}{\ifmmode {{\rm M}_{\odot}}\else M$_{\odot}$\fi}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\degree}{^{\circ}}

@@

\title{The Companion Mass Distribution to the ELM WD Sample}
\author{Jeff Andrews\altaffilmark{\colum}, Adrian M. Price-Whelan\altaffilmark{\colum}, Marcel Ag\"ueros\altaffilmark{\colum}}

@@

\begin{abstract}
Determining component masses of single-line spectroscopic binary stars
is impossible because of the unknown inclination angle. Even when the
mass of the bright star can be derived, only a lower limit can be placed
on the mass of the unseen companion. However since the probability of
any inclination angle scales with sin $i$, for a large enough, unbiased
sample, the companion mass distribution can be deconvolved from the
inclination angle distribution. We develop a Bayesian technique to handle this
deconvolution and place it within a Markov Chain Monte Carlo framework.
We then apply our method to the sample of 61 white dwarfs in the extremely low mass (ELM) white dwarf
survey. Modeling the companion mass
distribution as a Gaussian with a variable neutron star component, we
find the companion population is well modeled by a mean of {\bf 0.74} \Msun\ and a
standard deviation of {\bf 0.26} \Msun. Our model indicates a neutron star (NS)
fraction of {\bf 1.4$\substack{+7.0 \\ -1.4}$\%}. We make the posterior distribution from our simulations publicly available and provide a python script so that the probability of any future detected LMWD containing a NS can be determined.
\end{abstract}

@@

The dramatic increase in the number of spectroscopically confirmed white dwarfs (WDs) is seen in the most recently compiled sample from the Data Release 7 (DR7) in the Sloan Digital Sky Survey (SDSS), containing some 20,000 confirmed WDs \citet{kleinman13}. While the distribution of WD masses is strongly peaked at $0.6~\Msun$, a small but non-negligible number ($\approx$6\%) of low mass WDs (LMWD) is observed, with $M<0.4~\Msun$. In addition to SDSS, LMWDs have been observed in similar numbers elsewhere \citep[e.g. the Palomar Green Survey][]{liebert05}.

@@

In their seminal work, \citet{marsh95} discovered companions around five LMWDs in their sample of seven objects, confirming these theoretical expectations and cementing LMWDs as fruitful targets for binary searches. The Extremely Low Mass (ELM) WD Survey identifies LMWDs in SDSS for spectroscopic follow-up using a photometric constraint combined with a reduced proper motion constraint \citep{ELMI}. After their initial investigation found 11 of 12 LMWDs had close binary companions, further radial velocity follow-up has now brought the sample of ELM WDs to 61 systems, 55 of which have measured radial velocity amplitudes and orbital periods \citep{ELMII, ELMIII, ELMIV, ELMV}.

While the orbital periods and radial velocities indicated the companions were most likely WDs, due to the uncertain inclination angle, these LMWDs could have NS companions \citep{vLeeuwen07}. However this and other radio searches for pulsed emission, as well as searches for blackbody X-ray emission using {\it XMM-Newton} and {\it Chandra}, have all been unsuccessful \citep{agueros09a,agueros09b,kilic13}. Yet, such systems must exist since LMWDs are observed as companions around millisecond pulsars \citep{vKerkwijk96,callanan98,bassa06,antoniadis12}. Finding even one system would be extremely interesting; the NS binaries with WD companions bright enough to detect spectroscopic radial velocity variations have precisely determined NS masses, allowing for an understanding of the binary evolution processes forming these systems .

Selecting the most likely candidates for multiwavelength follow-up is crucial. For each WD in a single-line spectroscopic binary, spectroscopic observations provide the orbital period ($P_{\rm orb}$), WD mass ($M_1$), and projected orbital velocity ($K$). Using Kepler's third law, we can write the relation between these quantities and the unknown inclination angle as:

\begin{equation}
	\frac{M_2^3}{\left(M_1+M_2\right)^2} {\rm sin}^3 i = \frac{P_{\rm orb}}{2\pi G} K^3 = m_f \label{eq:massfunc}
\end{equation}

The righthand side of this equation is the well-known mass function ($M_f$). $M_2$ is minimized for an edge-on orbit, with $i = 90\degree$. For any individual system, deriving the likelihood of a $M_2$ for that system ($P(M_2)$) involves solving Equation \ref{eq:massfunc} for $M_2$ for different values for $i$, weighting the inclination angle by sin $i$. A basic calculation for the probability that the companion to any individual system is a NS can therefore be calculated:

\begin{equation}
P({\rm NS}) = \int_{0^{\circ}}^{i_{\rm NS}} {\rm sin}\ i\ {\rm d}i \label{eq:P_NS_approx}
\end{equation}

Where $i_{\rm NS}$ is the inclination angle from Equation \ref{eq:massfunc} when $M_2$ is the mass of a NS, typically $\approx1.35~\Msun$ \citep{thorsett99}. However, implicit in this calculation is that all companion masses are equally likely a priori. Clearly this is not the case as evidenced by \citep{kleinman13} who show the WD mass distribution in SDSS is strongly peaked around $0.6~\Msun$. Including this as a prior on the companion mass does not improve the situation since it is unlikely that the mass distribution of single WDs and WDs in binaries is the same. Equation \ref{eq:P_NS_approx} should therefore be used only as an approximation for the NS probability of a companion.

With 55 of the 61 systems with precise $M_1$, $P_{\rm orb}$, and $K$, the ELM sample is large enough to be statistically modeled. Furthermore, since the population was selected based on characteristics determined by the primary mass alone, at least with regard to the inclination angle and companion mass, the population is unbiased. In this work we develop a Bayesian model to determine the mass distribution of companions to ELM WDs. Our method is similar to that employed by \citet{ozel12} and \citet{kiziltan13} to describe the mass distribution of NSs in binaries. We focus on several specific scientific questions: 1.\ Can the population as a whole be modeled using a simple description of the companion masses? 2.\ How does the companion mass distribution compare to predictions from population synthesis simulations?  3.\ What is the rate of NS implied by our model? 4.\ Can we improve the likelihood of any individual system having a NS companion, taking into account the overall rate of NS's in the sample implied by our model?

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[angle=90,width=0.95\columnwidth]{Porb_M1.eps}
\caption{The $P_{\rm orb}$ - $M_1$ distribution of the ELM WD sample. The three eclipsing systems, with known $M_2$ are shown by the filled circles. Arrows indicate the masses of ELM WDs without detected radial velocity variation. Triangles indicate the positions of the known WD-NS binaries. From $P_{\rm orb}$ and $M_1$ alone, the two populations are indistinguishable.}
\end{center}
\end{figure}

@@

We construct a statistical model to derive constraints on a parametric model for the distribution of ELM WD companion masses, $p(M_2 \given \bs{\theta})$.\footnote{In what follows, vectors or sets of parameters or quantites are represented by bold symbols.} For each system, we assume we are given $K$, $T$, and $M_1$ and therefore know the value of the mass function, $m_f$ (Eq.~\ref{eq:massfunc}). We would like to find posterior constraints on the parameters, $\bs{\theta}$, that describe the distribution of companion masses, $p(M_2\given \bs{\theta})$, given all of the observed mass functions, $\bs{m}_f$, by deconvolving the distribution of mass functions from the unobserved inclination angle, $i$. Using Bayes' rule,
\begin{equation}
    p(\bs{\theta} \given m_f) = \frac{1}{\mathcal{Z}}~p(m_f \given \bs{\theta})~p(\bs{\theta}).
\end{equation}
where $p(m_f \given \bs{\theta})$ is the likelihood, $p(\bs{\theta})$ is the prior on parameters $\bs{\theta}$, and the evidence integral, $\mathcal{Z}$, is a constant that depends only on the data. The marginal likelihood, $p(m_f \given \bs{\theta})$, involves integrals over the latent variables $i$ and $M_2$,
\begin{align}
    p(m_f \given \bs{\theta}) &= \int_0^{\pi/2} di \int_{-\infty}^\infty dM_2  \nonumber \\
      & \qquad {} \times p(m_f \given M_1, M_2, i)~p(M_2 \given \bs{\theta})~p(i).
\end{align}
We neglect observational uncertainties in the mass function,\footnote{This is justified because the relative uncertainties in these quantities are small, $\sigma_x / x \sim 0.1$, however in future work we will extend this methodology to include uncertainties in all measured quantities.} and assume the inclination angles are isotropically distributed
\begin{align}
	p(m_f \given M_1, M_2, i) &= \delta \left[m_f - f(M_1, M_2, i) \right]\\
	f(M_1, M_2, i) &= \frac{(M_2 \sin i)^3}{(M_1 + M_2)^2}\\
	p(i) &= \sin i.
\end{align}
For now, we keep this general by not yet specifying a parametric form for the companion mass distribution, $p(M_2 \given \bs{\theta})$. With the above assumptions, the marginal likelihood integral is
\begin{align}
%    p(m_f \given \bs{\theta}) &= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta}) \int_0^{\pi/2} di \nonumber \\
%    & \qquad {} \times \delta \left[m_f - f(M_1, M_2, i)\right]~\sin i\\
    p(m_f \given \bs{\theta}) &= \int_{0}^\infty dM_2 ~p(M_2 \given \bs{\theta}) \int_0^{\pi/2} di  \nonumber \\
    & \qquad {} \times ~\sin i ~ \delta \left[g(M_1,M_2,i) \right]\label{eq:delta}
\end{align}
where
\begin{equation}
	g(M_1,M_2,i) = \frac{M_2^3}{(M_1+M_2)^2}\sin^3 i - m_f.
\end{equation}
The inner integral (over $i$) has the form
\begin{equation}
    \int dx~F(x)~\delta \left[ G(x) \right] = \sum_j \frac{F(x^*_j)}{|G'(x^*_j)|}
\end{equation}
where the sum is over the roots, $x^*_j$, of the function $G(x)$. The root, $i^*$, and derivative of the argument of the delta function in Equation~\ref{eq:delta} are 
\begin{align}
%	\sin i^* &= \alpha\\
%	\alpha &= \frac{(m_f(M_1+M_2)^2)^{1/3}}{M_2}\\
	\sin i^* &= \frac{ \left[m_f(M_1+M_2)^2 \right]^{1/3}}{M_2}\\
%	\frac{\partial g}{\partial i} &= \frac{3M_2^3}{(M_1+M_2)^2}\sin^2 i \cos i\\
%	\frac{\partial g}{\partial i}\bigg\rvert_{i^*} &= \frac{3M_2^3}{(M_1+M_2)^2} \alpha^2 \sqrt{1 - \alpha^2}
	\frac{\partial g}{\partial i}\bigg\rvert_{i^*} &= \frac{3M_2^3}{(M_1+M_2)^2} \sin^2 i^* \sqrt{1 - \sin^2 i^*}
\end{align}
With this in mind, we may write the marginal likelihood as
\begin{align}
	p(m_f \given \bs{\theta}) &= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta})~\frac{\sin i^*}{g'(M_1,M_2,i^*)}\\
%	&= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta})~\alpha\left[\frac{3M_2^3}{(M_1+M_2)^2} \alpha^2 \sqrt{1 - \alpha^2}\right]^{-1}\\
	&= \int_{M_{2,{\rm min}}}^\infty dM_2~p(M_2 \given \bs{\theta})~h(M_2, m_f, M_1) \label{eq:fullm2}
\end{align}

where the bottom bound in the integral in Equation~\ref{eq:fullm2} is set by the minimum companion mass for which the integrand is real, $M_{2,{\rm min}}$ (determined by setting $i=90\degree$ in Equation~\ref{eq:massfunc} and solving for $M_2$), and

\begin{equation}
h(M_2, m_f, M_1) = \frac{(M_1+M_2)^{4/3}}{3\ m_f^{1/3}M_2\sqrt{M_2^2 - \left[ m_f(M_1+M_2)^2 \right]^{2/3}}}
\end{equation}

\section{Our Model} \label{sec:experiments}

We must now choose a functional form for the companion mass distribution,  $p(M_2\given \bs{\theta})$. For all experiments -- including that with the real data -- we use a two component Gaussian mixture model to fit the data. However, in tests below, we generate test data with a variety of mixture component forms. We truncate the distributions using physically motivated bounds so that the white dwarf component is restricted to the range $M_2\in (0.2,1.4)~\Msun$ and the neutron star component is restricted to the range $M_2\in (1,2)~\Msun$. The likelihood with a Gaussian mixture model is
\begin{align}
%	h(M_2, m_f, M_1) &= \frac{(M_1+M_2)^{4/3}}{3m_f^{1/3}M_2\sqrt{M_2^2 - \left[ m_f(M_1+M_2)^2 \right]^{2/3}}}\\
%	p(m_f \given \bs{\theta}) &= \int_{0}^\infty dM_2~\left[ (1-f_{\rm NS})p_{\rm WD} + f_{\rm NS}p_{\rm NS} \right]~h(M_2, m_f, M_1)\\
	p(M_2 \given \bs{\theta}) &= \left[ (1-f_{\rm NS})p_{\rm WD} + f_{\rm NS}p_{\rm NS} \right] 
\end{align}

where

\begin{equation}
	p_{\rm WD} = \mathcal{N}(M_2 \given \mu_{\rm WD}, \sigma^2_{\rm WD});\,\left[ 0.2 < \frac{M_2}{\Msun} < 1.4 \right] 
\end{equation}

and

\begin{equation}
	p_{\rm NS} = \mathcal{N}(M_2 \given \mu_{\rm NS}, \sigma^2_{\rm NS});\, \left[1.4 < \frac{M_2}{\Msun} < 2 \right].
\end{equation}

$\mathcal{N}$ is the (truncated, but properly normalized) normal distribution with mean $\mu$ and variance $\sigma^2$ and the distributions are limited to the ranges specified. In this work, we fix the mean and variance of the neutron star distribution to
\begin{align}
	\mu_{\rm NS} &= 1.4~\Msun\\
	\sigma^2_{\rm NS} &= (0.1~\Msun)^2
\end{align}
based on indications that NSs in certain binaries may be somewhat more massive than the canonical NS mass of 1.35 \Msun \citep{kiziltan13,smedley14}.

%The probability of any particular system having a NS companion $P({\rm NS})$ is then:

%\begin{equation}
%P({\rm NS}) = \int_{0}^\infty dM_2~ f_{\rm NS}~ \mathcal{N}(M_2 \given \mu_{\rm NS}, \sigma^2_{\rm NS})~ h(M_2, m_f, M_1) \label{eq:P_NS_full}
%\end{equation}

Our companion mass model parameters are then $\bs{\theta} = (\mu_{\rm WD}, \sigma_{\rm WD}, f_{\rm NS})$. We choose standard priors on these parameters. For the mean of the white dwarf component, we use a uniform distribution from $0.2-1.4~\Msun$. We use a (scale-invariant) logarithmic prior for the variance of the white dwarf component over the range $0.05-2.0~\Msun$. We use a uniform distribution over the dimensionless neutron star fraction from $0-1$. The model parameters are summarized in Table~\ref{tbl:parameters}.

@@

In our model, we use an ensemble Markov Chain Monte Carlo algorithm \citep{goodman10} to draw samples from the posterior distribution, $p(\mu_{\rm WD}, \sigma_{\rm WD}, f_{\rm NS} \given \bs{m}_f, \bs{M}_1)$. The algorithm is implemented in the \texttt{Python} programming language \citep{foremanmackey13} and uses an ensemble of individual ``walkers'' to naturally adapt to the geometry of the parameter-space being explored. We run the walkers for an initial, burn-in period of 1000 steps starting from randomly drawn initial conditions (sampled from the priors in Table~\ref{tbl:parameters}). We then re-initialized the walkers from their positions at the end of this run, throw out the burn-in samples, and run again for 1000 steps. We compute the autocorrelation time for each parameter to confirm they have converged.

\subsection{Test Cases} \label{sec:tests}
To test the performance of this Gaussian mixture model, we apply it to three separate sets of mock data (described in detail below), each with 100 ``observed'' systems. These data are generated by randomly drawing the companion mass, $M_2$, from a particular distribution, then using a randomly drawn inclination angle and primary mass, $M_1$, to compute the mass function, $m_f$. We draw primary masses from a uniform distribution, $M_1 \sim \mathcal{U}(0.2,0.4)~\Msun$. We apply the same Gaussian mixture model to all test datasets to infer the parameters of the white dwarf mixture component and neutron star fraction.

\begin{figure*}[h!]
\begin{center}
\includegraphics[width=0.95\textwidth]{many-panel.pdf}
\caption{The model results of our three tests described in Section~\ref{sec:tests}. The left-most panel shows a histogram of the $M_2$ drawn from each of our test distributions. Overplotted as a solid line is posterior distribution using the maximum a posteriori parameter values, resulting from our model. The second panel shows the posterior distributions for $\mu_{\rm WD}$ and $\sigma_{\rm WD}$. The dashed lines in the top two panels show the true values that the sample systems were drawn from. Since the third model was not drawn from a Gaussian distribution, there is no true parameter set. The third panel shows the convolved posterior histogram for $f_{\rm NS}$, with the dashed lines in the bottom two panels showing the true value of 10\%. Finally, the rightmost panels show the NS probability for individual systems as a function of $m_f$. The error bars in $P_{\rm NS}$ enclose the interquartile range.}
\label{fig:tests}
\end{center}
\end{figure*}

\subsubsection{Test 1: Single Gaussian (white dwarf)} \label{sec:exp1}

As a first test, we generate $M_2$'s by drawing from a single, truncated Gaussian with mean, variance, and bounds
\begin{align}
	\mu_{\rm WD, true} &= 0.7~\Msun\\
	\sigma^2_{\rm WD, true} &= (0.2~\Msun)^2\\
	0.2 < &\frac{M_2}{\Msun} < 1.4.
\end{align}
This is equivalent to setting the neutron star fraction to $f_{\rm NS} = 0$. The distribution of mock observations of the mass function, $m_f$, are shown by the top-left panel of Figure~\ref{fig:tests}. We use MCMC (see description above) to ...

>>>>

% commands
\newcommand{\given}{\,|\,}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transpose}[1]{{#1}^{\mathsf{T}}}
\newcommand{\inverse}[1]{{#1}^{-1}}
\newcommand{\Msun}{\ifmmode {{\rm M}_{\odot}}\else M$_{\odot}$\fi}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\degree}{^{\circ}}
\newcommand{\eqn}{Equation~}

% Symbols
\newcommand{\period}{T_{\rm orb}}
\newcommand{\mf}{M_f}
\newcommand{\wdupper}{1.44}

@@

\title{The Companion Mass Distribution of extremely low-mass white dwarfs}
\author{Jeff Andrews\altaffilmark{\colum}, Adrian M. Price-Whelan\altaffilmark{\colum}, Marcel Ag\"ueros\altaffilmark{\colum}}

@@

\begin{abstract}
Measuring the masses of companions to single-line spectroscopic binary stars
is (in general) not possible because of the unknown orbital plane inclination angles. 
Even when the mass of the visible star can be measured, only a lower limit can be placed
on the mass of the unseen companion. However, since the systems should be isotropically
distributed in the inclination angle, for a large enough, unbiased sample, the 
companion mass distribution can be deconvolved from the distributions of observables.
In this work, we construct a hierarchical probabilistic model to infer properties of unseen 
companion stars given kinematic measurements of the primary. We use a mixture of 
(two) Gaussians to model the true companion mass distribution, meant to represent 
populations of white dwarfs (WD) and neutron stars (NS). We run this model on mock data generated
both with and without the same functional form as the fitting model in order to test the
robustness of our inferences. We then apply our model to 55 of the 61 white dwarfs in 
the extremely low mass (ELM) WD survey. We find maximum a posteriori (MAP) parameters for
the WD companion population of $\mu_{\rm WD} = 0.73~\Msun$ and $\sigma_{\rm WD} = 0.26~\Msun$ 
with a NS fraction consistent with zero, $f_{\rm NS} = 0$. However, the marginal posterior 
distribution over NS fraction has a significant tail up to $f_{\rm NS} \sim 0.12$.
We make samples from the posterior distribution publicly available and provide a  
script for computing the probability that a ... in the context of our model.
\end{abstract}

@@

The dramatic increase in the number of spectroscopically confirmed white dwarfs (WDs) is seen in the most recently compiled sample from Data Release 7 (DR7) of the Sloan Digital Sky Survey (SDSS), which contains $\sim$20,000 confirmed WDs \citep{kleinman13}. While the distribution of WD masses is strongly peaked at $0.6~\Msun$, a small but non-negligible number ($\sim$6\%) of low mass WDs (LMWD) is observed with $M<0.4~\Msun$. In addition to SDSS, LMWDs have been observed in similar numbers elsewhere \citep[e.g., the Palomar Green Survey;][]{liebert05}.

@@

In their seminal work, \citet{marsh95} discovered companions around five LMWDs in their sample of seven objects, confirming these theoretical expectations and cementing LMWDs as fruitful targets for binary searches. The Extremely Low Mass (ELM) WD Survey identifies LMWDs in SDSS for spectroscopic follow-up using combined constraints on photometry and reduced proper motions \citep{ELMI}. After their initial investigation found 11 of 12 LMWDs had close binary companions, further radial velocity follow-up has now brought the sample of ELM WDs to 61 systems, 55 of which have measured radial velocity amplitudes and orbital periods \citep{ELMII, ELMIII, ELMIV, ELMV}.

While the orbital periods and radial velocities indicated the companions were most likely WDs, due to the unmeasured inclination angle, these LMWDs could have NS companions \citep{vLeeuwen07}. However, this and other radio searches for pulsed emission, as well as searches for blackbody X-ray emission using {\it XMM-Newton} and {\it Chandra}, have all been unsuccessful \citep{agueros09a,agueros09b,kilic13}. Yet, such systems must exist since LMWDs are observed as companions around millisecond pulsars \citep{vKerkwijk96,callanan98,bassa06,antoniadis12}. Finding even one system would be extremely interesting: the NS binaries with WD companions bright enough to detect spectroscopic radial velocity variations have precisely determined NS masses which provide useful insights into the binary evolution processes that form these systems .

Selecting the most likely candidates for multiwavelength follow-up is crucial. For each WD in a single-line spectroscopic binary, spectroscopic observations provide the orbital period ($\period$), WD mass ($M_1$), and projected orbital velocity ($K=v \sin i$). Using Kepler's third law, we can write the relation between these quantities and the unknown inclination angle as:
\begin{equation}
	\frac{(M_2 \sin i)^3}{\left(M_1+M_2\right)^2} = \frac{\period}{2\pi G} K^3 = \mf \label{eq:massfunc}
\end{equation}
The righthand side of this equation is the well-known mass function ($\mf$). $M_2$ is minimized for an edge-on orbit, with $i = 90\degree$.

55 of the 61 systems have precise $M_1$, $\period$, and $K$ measured; we refer to this as the ELM sample. Since the population was selected based on characteristics determined by the primary mass alone, at least with regard to the inclination angle and companion mass, the population is unbiased. In this work, we develop a probabilistic model to infer parameters of an assumed form for the mass distribution of companions to ELM WDs. Our method is similar to that employed by \citet{ozel12} and \citet{kiziltan13} to describe the mass distribution of NSs in binaries. We focus on several specific scientific questions: 1.\ Can the population as a whole be modeled using a simple description of the companion masses? 2.\ How does the companion mass distribution compare to predictions from population synthesis simulations?  3.\ What is the rate of NS-LMWD binaries implied by our model? 4.\ {\bf [Jeff: I propose nixing this question... ]} Can we improve the likelihood of any individual system having a NS companion, taking into account the overall rate of NS's in the sample implied by our model?

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[angle=90,width=0.95\columnwidth]{Porb_M1.eps}
\caption{The $\period$ - $M_1$ distribution of the ELM WD sample. The three eclipsing systems, with known $M_2$ are shown by the filled circles. Arrows indicate the masses of ELM WDs without detected radial velocity variation. Triangles indicate the positions of the known WD-NS binaries. From $\period$ and $M_1$ alone, the two populations are indistinguishable.}
\end{center}
\end{figure}

@@

We construct a statistical model to derive constraints on a parametric model for the distribution of ELM WD companion masses, $p(M_2 \given \bs{\theta})$.\footnote{In what follows, vectors or sets of parameters or quantites are represented by bold symbols.} For each system, we assume we are given $K$, $T$, and $M_1$ and therefore know the value of the mass function, $\mf$ (\eqn~\ref{eq:massfunc}). We would like to derive posterior constraints on the parameters, $\bs{\theta}$, that describe the distribution of companion masses, $p(M_2\given \bs{\theta})$, given all of the observed mass functions, $\bs{m}_f$, by deconvolving the distribution of mass functions from the unobserved inclination angle, $i$. Using Bayes' rule,
\begin{equation}
    p(\bs{\theta} \given \mf) = \frac{1}{\mathcal{Z}}~p(\mf \given \bs{\theta})~p(\bs{\theta}).
\end{equation}
where $p(\mf \given \bs{\theta})$ is the likelihood, $p(\bs{\theta})$ is the prior on parameters $\bs{\theta}$, and the evidence integral, $\mathcal{Z}$, is a constant that depends only on the data. The marginal likelihood, $p(\mf \given \bs{\theta})$, involves integrals over the latent variables $i$ and $M_2$,
\begin{align}
    p(\mf \given \bs{\theta}) &= \int_0^{\pi/2} di \int_{-\infty}^\infty dM_2  \nonumber \\
      & \qquad {} \times p(\mf \given M_1, M_2, i)~p(M_2 \given \bs{\theta})~p(i).
\end{align}
We neglect observational uncertainties in the mass function,\footnote{This is justified because the relative uncertainties in these quantities are small, $\sigma_x / x \sim 0.05-0.1$, however in future work we will extend this methodology to include uncertainties in all measured quantities.} and assume the inclination angles are isotropically distributed
\begin{align}
	p(\mf \given M_1, M_2, i) &= \delta \left[\mf - f(M_1, M_2, i) \right]\\
	f(M_1, M_2, i) &= \frac{(M_2 \sin i)^3}{(M_1 + M_2)^2}\\
	p(i) &= \sin i.
\end{align}
For now, we keep this general by not yet specifying a parametric form for the companion mass distribution, $p(M_2 \given \bs{\theta})$. With the above assumptions, the marginal likelihood integral is
\begin{align}
%    p(\mf \given \bs{\theta}) &= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta}) \int_0^{\pi/2} di \nonumber \\
%    & \qquad {} \times \delta \left[\mf - f(M_1, M_2, i)\right]~\sin i\\
    p(\mf \given \bs{\theta}) &= \int_{0}^\infty dM_2 ~p(M_2 \given \bs{\theta}) \int_0^{\pi/2} di  \nonumber \\
    & \qquad {} \times ~\sin i ~ \delta \left[g(M_1,M_2,i) \right]\label{eq:delta}
\end{align}
where
\begin{equation}
	g(M_1,M_2,i) = \frac{M_2^3}{(M_1+M_2)^2}\sin^3 i - \mf.
\end{equation}
The inner integral (over $i$) has the form
\begin{equation}
    \int dx~F(x)~\delta \left[ G(x) \right] = \sum_j \frac{F(x^*_j)}{|G'(x^*_j)|}
\end{equation}
where the sum is over the roots, $x^*_j$, of the function $G(x)$. The root, $i^*$, and derivative of the argument of the delta function in \eqn\ref{eq:delta} are 
\begin{align}
%	\sin i^* &= \alpha\\
%	\alpha &= \frac{(\mf(M_1+M_2)^2)^{1/3}}{M_2}\\
	\sin i^* &= \frac{ \left[\mf(M_1+M_2)^2 \right]^{1/3}}{M_2}\\
%	\frac{\partial g}{\partial i} &= \frac{3M_2^3}{(M_1+M_2)^2}\sin^2 i \cos i\\
%	\frac{\partial g}{\partial i}\bigg\rvert_{i^*} &= \frac{3M_2^3}{(M_1+M_2)^2} \alpha^2 \sqrt{1 - \alpha^2}
	\frac{\partial g}{\partial i}\bigg\rvert_{i^*} &= \frac{3M_2^3}{(M_1+M_2)^2} \sin^2 i^* \sqrt{1 - \sin^2 i^*}
\end{align}
With this in mind, we may write the marginal likelihood as
\begin{align}
	p(\mf \given \bs{\theta}) &= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta})~\sin i^* \left(\frac{\partial g}{\partial i}\bigg\rvert_{i^*}\right)^{-1}\\
%	&= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta})~\alpha\left[\frac{3M_2^3}{(M_1+M_2)^2} \alpha^2 \sqrt{1 - \alpha^2}\right]^{-1}\\
	&= \int_{M_{2,{\rm min}}}^\infty dM_2~p(M_2 \given \bs{\theta})~h(M_2, \mf, M_1) \label{eq:fullm2}
\end{align}

where the bottom bound in the integral in \eqn\ref{eq:fullm2} is set by the minimum companion mass for which the integrand is real, $M_{2,{\rm min}}$ (determined by setting $i=90\degree$ in \eqn\ref{eq:massfunc} and solving for $M_2$), and

\begin{equation}
h(M_2, \mf, M_1) = \frac{(M_1+M_2)^{4/3}}{3\ \mf^{1/3}M_2\sqrt{M_2^2 - \left[ \mf(M_1+M_2)^2 \right]^{2/3}}}
\end{equation}

\subsection{Our Model} \label{sec:experiments}

We must now choose a functional form for the companion mass distribution,  $p(M_2\given \bs{\theta})$. For all experiments -- including that with the real data -- we use a two component Gaussian mixture model to fit the data. However, in tests below, we generate test data with a variety of mixture component forms. We truncate the distributions using physically motivated bounds so that the white dwarf component is restricted to the range $M_2\in (0.2,\wdupper)~\Msun$ and the neutron star component is restricted to the range $M_2\in (1.3,2)~\Msun$,
\begin{align}
%	h(M_2, \mf, M_1) &= \frac{(M_1+M_2)^{4/3}}{3\mf^{1/3}M_2\sqrt{M_2^2 - \left[ \mf(M_1+M_2)^2 \right]^{2/3}}}\\
%	p(\mf \given \bs{\theta}) &= \int_{0}^\infty dM_2~\left[ (1-f_{\rm NS})p_{\rm WD} + f_{\rm NS}p_{\rm NS} \right]~h(M_2, \mf, M_1)\\
	p(M_2 \given \bs{\theta}) &= \left[ (1-f_{\rm NS})p_{\rm WD} + f_{\rm NS}p_{\rm NS} \right] 
\end{align}
where 
\begin{align}
	p_{\rm WD} &= \mathcal{N}(M_2 \given \mu_{\rm WD}, \sigma^2_{\rm WD});\,\left[ 0.2 < \frac{M_2}{\Msun} < \wdupper \right] \\
	p_{\rm NS} &= \mathcal{N}(M_2 \given \mu_{\rm NS}, \sigma^2_{\rm NS});\, \left[1.3 < \frac{M_2}{\Msun} < 2 \right].
\end{align}
and $\mathcal{N}$ is the (truncated, but properly normalized) normal distribution with mean $\mu$ and variance $\sigma^2$ and the distributions are limited to the ranges specified. In this work, we fix the mean and variance of the neutron star distribution to
\begin{align}
	\mu_{\rm NS} &= 1.4~\Msun\\
	\sigma^2_{\rm NS} &= (0.05~\Msun)^2
\end{align}
based on indications that NSs in certain binaries may be somewhat more massive than the canonical NS mass of 1.35 \Msun \citep{kiziltan13,smedley14}.

%The probability of any particular system having a NS companion $P_{\rm NS}$ is then:

%\begin{equation}
%P_{\rm NS} = \int_{0}^\infty dM_2~ f_{\rm NS}~ \mathcal{N}(M_2 \given \mu_{\rm NS}, \sigma^2_{\rm NS})~ h(M_2, \mf, M_1) \label{eq:P_NS_full}
%\end{equation}

Our companion mass model parameters are then $\bs{\theta} = (\mu_{\rm WD}, \sigma_{\rm WD}, f_{\rm NS})$. We choose standard priors on these parameters: for the mean of the white dwarf component, we use a uniform distribution from $0.2-1.4~\Msun$; we use a (scale-invariant) logarithmic prior for the variance of the white dwarf component over the range $0.05-2~\Msun$; finally, we use a uniform prior over the dimensionless neutron star fraction from $0-1$. The model parameters are summarized in Table~\ref{tbl:parameters}.

@@

We use an ensemble Markov Chain Monte Carlo algorithm \citep{goodman10} to draw samples from the posterior distribution, $p(\mu_{\rm WD}, \sigma_{\rm WD}, f_{\rm NS} \given \bs{m}_f, \bs{M}_1)$. The algorithm is implemented in the \texttt{Python} programming language \citep{foremanmackey13} and uses an ensemble of individual ``walkers'' to naturally adapt to the geometry of the parameter-space being explored. We run the walkers for an initial, burn-in period of 1000 steps starting from randomly drawn initial conditions (sampled from the priors in Table~\ref{tbl:parameters}). We then re-initialized the walkers from their positions at the end of this run, throw out the burn-in samples, and run again for 1000 steps. We compute the autocorrelation time for each parameter to confirm they have converged.

\section{Test Cases} \label{sec:tests}
To test the performance of this Gaussian mixture model, we apply it to three separate sets of mock data (described in detail below), each with 100 ``observed'' systems. These data are generated by randomly drawing the companion mass, $M_2$, from a particular distribution, then using a randomly drawn inclination angle and primary mass, $M_1$, to compute the mass function, $\mf$. We draw primary masses from a uniform distribution, $M_1 \sim \mathcal{U}(0.2,0.4)~\Msun$. We apply the same Gaussian mixture model to all test datasets to infer the parameters of the white dwarf mixture component and neutron star fraction.

\begin{figure*}[h!]
\begin{center}
\includegraphics[width=0.95\textwidth]{many-panel.pdf}
\caption{The model results of our three tests described in Section~\ref{sec:tests}. The left-most panel shows a histogram of the $M_2$ drawn from each of our test distributions. Overplotted as a solid line is posterior distribution using the maximum a posteriori parameter values, resulting from our model. The second panel shows the posterior distributions for $\mu_{\rm WD}$ and $\sigma_{\rm WD}$. The dashed lines in the top two panels show the true values that the sample systems were drawn from. Since the third model was not drawn from a Gaussian distribution, there is no true parameter set. The third panel shows the convolved posterior histogram for $f_{\rm NS}$, with the dashed lines in the bottom two panels showing the true value of 10\%. Finally, the rightmost panels show the NS probability for individual systems as a function of $\mf$. The error bars in $P_{\rm NS}$ enclose the interquartile range.}
\label{fig:tests}
\end{center}
\end{figure*}

\subsection{Test 1: Single Gaussian (white dwarf)} \label{sec:exp1}

As a first test, we generate $M_2$'s by drawing from a single, truncated Gaussian with mean, variance, and bounds
\begin{align}
	\mu_{\rm WD, true} &= 0.7~\Msun\\
	\sigma^2_{\rm WD, true} &= (0.2~\Msun)^2\\
	0.2 < &\frac{M_2}{\Msun} < \wdupper.
\end{align}
This is equivalent to setting the neutron star fraction to $f_{\rm NS} = 0$.
